Visualizing Algebraic Surfaces

### 2016-04-22 project/article/src/article.tex
typing out a s-storm

\usepackage[backend=biber, style=alphabetic]{biblatex} 
\addbibresource{../bib/mat2000.bib}

\usepackage[demo]{graphicx} 
\usepackage[]{caption} 
\usepackage[]{subcaption}

@@

\theoremstyle{definition}
\newtheorem{defn}{Definition}

\newcommand{\C}{\ensuremath{\mathbb{C}}}
\newcommand{\proj}{\ensuremath{\mathbb{P}}}

@@

    \maketitle
    \tableofcontents

@@

    \begin{figure}[h]
        \centering
        \begin{subfigure}{0.5\textwidth}
            \centering
            \includegraphics[width=0.35\linewidth]{foo}
            \caption{$g(x, y) = y^2 - x^3 = 0$}
        \end{subfigure}%
        \begin{subfigure}{0.5\textwidth}
            \centering
            \includegraphics[width=0.35\linewidth]{bar}
            \caption{$g(x, y) = y^2 - x^3 + x^2 = 0$}
        \end{subfigure} 
        \caption{Examples of curves that are \emph{not} graphs given by their
        functions $g(x, y)$}.}
        \label{fig:curves_not_graphs}
    \end{figure}

@@

    The surfaces considered so far are over the real numbers, i.e, the
    coefficients $a_{ijk}$ are real numbers. We can instead work over the
    complex numbers where the surfaces are given by an equivalent equation, but
    where the coefficients now are complex numbers\footnote{We will come back
    to why we make this transition when we talk about sets being algebraically
    closed or not.}:
    \begin{equation}
        \left\{ (x, y, z) \in \C \mid h(x, y, z) = 0 \right\}.
    \end{equation}
    Again, doing the same trick (one-trick ponies) we can now define $w = h(x,
    y, z)$ and look at the equations $w - h(x, y, z) = 0$. Now, we are stuck
    with an equation in four variables. With only three degrees of freedom to
    work with when visualizing these objects we encounter an obstacle. How do
    we know what these surfaces look like? This brings us to the world of
    \emph{projective geometry}.

@@

    The way we visualize these surfaces in four variables is by \emph{fixing}
    one of the variables. This is called \emph{homogenizing} the polynomial
    $h(x, y, z)$. We first want to introduce some definitions.

@@

    \begin{defn}[The projective space of dimension $n$]
        Let $\sim$ be an equivalence relation on $\C^{n+1} \setminus \left\{ 0
        \right\}$ defined by
        \begin{equation}
            \notag
            \left( x_1, \ldots, x_{n+1} \right) \sim \left( x_1', \ldots,
            \x_{n+1}' \right) \iff \left( x_1', \ldots, x_{n+1}' \right) =
            \lambda \left( x_1, \ldots, x_{n+1} \right)
        \end{equation}
        where $\lambda$ is some non-zero complex number. We then denote the
        equivalence class of $\left( x_1, \ldots, x_{n+1} \right)$ by $\left[
        x_1 : \ldots : x_{n+1} \right]$ and define the \emph{projective space
        of dimension $n$} as the set
        \begin{equation}
            \notag
            \proj^n{\C} = \left\{ \left[ x_1 : \ldots : x_{n+1} \right] \mid
            \left( x_1, \ldots, x_{n+1} \right) \in \C^{n+1} \setminus \left\{
            0 \right\} \right\}.
        \end{equation}
    \end{defn}

\printbibliography

\end{document}

>>>>

\usepackage[backend=biber, style=alphabetic]{biblatex} 
\addbibresource{../bib/mat2000.bib}
\usepackage[]{url} 
\urlstyle{sf}

\usepackage[]{graphicx} 
\usepackage[]{wrapfig} 
\usepackage[]{caption} 
\usepackage[]{subcaption}

@@

\theoremstyle{definition}
\newtheorem{defn}{Definition}
\newtheorem{exmpl}{Example}

\theoremstyle{plain}
\newtheorem{lma}{Lemma}
\newtheorem{crl}{Corollary}

\newcommand{\C}{\ensuremath{\mathbb{C}}}
\newcommand{\proj}{\ensuremath{\mathbb{P}}}
\renewcommand{\hom}{\ensuremath{h_{\mathrm{hom}}}}

@@

    \begin{titlepage}
    \maketitle

    \begin{figure}[htb]
        \centering
        \includegraphics[width=0.8\linewidth]{../pictures/labs_septic.png}
    \end{figure}
    \end{titlepage}
    \tableofcontents

@@

    \begin{figure}[h]
        \centering
        \begin{subfigure}{0.5\textwidth}
            \centering
            \includegraphics[width=1.0\linewidth]{../pictures/cusp.pdf}
            \caption{$g(x, y) = y^2 - x^3 = 0$}
        \end{subfigure}%
        \begin{subfigure}{0.5\textwidth}
            \centering
            \includegraphics[width=1.0\linewidth]{../pictures/double_point.pdf}
            \caption{$g(x, y) = y^2 - x^3 + x^2 = 0$}
        \end{subfigure} 
        \caption{Examples of curves that are \emph{not} graphs given by their
        functions $g(x, y)$}.}
        \label{fig:curves_not_graphs}
    \end{figure}

@@

    The surfaces considered so far are over the real numbers, i.e, the
    coefficients $a_{ijk}$ are real numbers. We can instead work over the
    complex numbers where the surfaces are given by an equivalent equation, but
    where the coefficients now are complex numbers\footnote{We will come back
    to why we make this transition when we talk about sets being algebraically
    closed or not.}:
    \begin{equation}
        \left\{ (x, y, z) \in \C \mid h(x, y, z) = 0 \right\}.
    \end{equation}
    Again, doing the same trick (one-trick ponies) we can now define $w = h(x,
    y, z)$ and look at the equations $w - h(x, y, z) = 0$ in the $(x, y, z,
    w)$-space. Now, we are stuck with an equation in four variables. With only
    three degrees of freedom to work with when visualizing these objects we
    encounter an obstacle. How do we know what these surfaces look like? This
    brings us to the world of \emph{projective geometry}.

@@

    The way we work these surfaces in four variables is by introducing a new
    variable making the polynomial terms all have the same degree.  This is
    called \emph{homogenizing} the polynomial $h(x, y, z)$. We first want to
    introduce some definitions.

@@

    \begin{defn}[The projective space of dimension $n$]
        Let $\sim$ be an equivalence relation on $\C^{n+1} \setminus \left\{ 0
        \right\}$ defined by
        \begin{equation}
            \notag
            \left( x_1, \ldots, x_{n+1} \right) \sim \left( x_1', \ldots,
            x_{n+1}' \right) \iff \left( x_1', \ldots, x_{n+1}' \right) =
            \lambda \left( x_1, \ldots, x_{n+1} \right)
        \end{equation}
        where $\lambda$ is some non-zero complex number. We then denote the
        equivalence class of $\left( x_1, \ldots, x_{n+1} \right)$ by $\left[
        x_1 : \ldots : x_{n+1} \right]$ and define the \emph{projective space
        of dimension $n$} as the set
        \begin{equation}
            \notag
            \proj^n{\C} = \left\{ \left[ x_1 : \ldots : x_{n+1} \right] \mid
            \left( x_1, \ldots, x_{n+1} \right) \in \C^{n+1} \setminus \left\{
            0 \right\} \right\}.
        \end{equation}
    \end{defn}

    \section{Algebraic Surfaces}
    \label{sec:algebraic_surfaces}

    We are now equipped with what we need to again look at our algebraic
    surfaces defined by $h(x, y, z) = 0$. The solutions to this equation are
    points in $\C^4$. We homogenize this polynomial by introducing the variable
    $w$ and evaluating $h$ at the points $x \to x/w$, $y \to y/w$ and $z \to
    z/w$ and then multiply by $w^d$ where $d$ is the degree of the polynomial
    $h(x, y, z)$.

    \begin{defn}[Homogenized polynomial]
        Let $h(x, y, z)$ be a polynomial in $x, y, z$ of degree $d$. 
        The homogenized polynomial $\hom(w, x, y, z)$ is given by
        \begin{equation}
            \notag
            \hom(w, x, y, z) = w^d h \left( \tfrac{x}{w}, \tfrac{y}{w}, \tfrac{z}{w} \right).
        \end{equation}
        It follows that each non-zero term in this polynomial has the same
        degree.
    \end{defn}
    The property that each non-zero term has the same degree in a homogeneous
    polynomial gives rise to a very important property which relates directly
    to the projective geometry introduced above.\footnote{This following
    property can be seen directly by visualizing the homogenous polynomial
    formed by a polynomial $g(x, y)$ and noticing that the figure does not
    change when zooming in or out.}

    \begin{crl}
        If $\hom$ is an homogeneous polynomial formed by homogenizing the
        polynomial $h(x, y, z)$ of degree $d$, then
        \begin{equation}
            \notag
            \hom(\lambda w, \lambda x, \lambda y, \lambda z) = \lambda^d \hom(w, x, y, z).
        \end{equation}
    \end{crl}
    \begin{proof}
        We simply compute the left hand side:
        \begin{align*}
            \notag
            \hom(\lambda w, \lambda x, \lambda y, \lambda z) &= (\lambda w)^d h
            \left( \tfrac{\lambda x}{\lambda w}, \tfrac{\lambda x}{\lambda w},
            \tfrac{\lambda x}{\lambda w}\right) \\ &= \lambda^d \hom(w, x, y,
            z).
        \end{align*}
    \end{proof}
    Since we are working with homogenous coordinates, the condition $\hom(w, x,
    y, z) = 0$ is well defined on the equivalence class discussed above,
    strictly due to this first corollary.

    Now, if we can homogenize a polynomial, we can certainly dehomogenize it.
    Given a homogeneous polynomial $\hom(w, x, y, z)$ we can dehomogenize the
    polynomial by \emph{fixing} one of the variables.

    \subsection{Initial examples of algebraic surfaces}
    \label{sub:initial_examples_of_algebraic_surfaces}

    Although mentioned before, we repeat the definition of an algebraic surface.
    \begin{defn}[Algebraic surface]
        A surface given by the equation $h(x, y, z) = 0$ is called \emph{algebraic}
        if $h(x, y, z)$ is a polynomial in the variables $x, y, z$.
        Mathematically,
        \begin{equation}
            \notag
            h(x, y, z) = \sum^{}_{i, j, k} a_{ijk}x^iy^jz^k.
        \end{equation}
    \end{defn}

    We are also interested in classifying the surfaces according to their
    \emph{degree}. 
    \begin{defn}[The degree of a surface]
        Given a surface defined by the equation $h(x, y, z) = 0$ we define the
        \emph{degree}, denoted $d$, of the surface to be the largest sum of
        powers in the terms of the polynomial. Mathematically,
        \begin{equation}
            \notag
            d = \max\{i + j + k\}
        \end{equation}
        where $i, j, k$ are the powers of $x, y$ and $z$ in the terms.
    \end{defn}
    Now, what do these surfaces actually look like? We give some examples that
    is going to illustrate the above definition.

    \begin{exmpl}
        Let $h$ be a polynomial in $x, y, z$ given by
        \begin{equation}
            \notag h(x, y, z) = x^2 + y^2 - z^2. 
        \end{equation}
        We then look at the set of solutions to the equation $h(x, y, z) = 0$.
        Plotting these yields a double-cone shaped figure. We see there is an
        interesting quirk at the point $(0, 0, 0)$ where the surface seems to
        pass through itself, formally we say that the surface exhibits a
        \emph{singularity} at the point $(0, 0, 0)$, but we will come back to
        that later.
        \begin{figure}[h!]
            \centering
            \includegraphics[width=0.3\linewidth]{../pictures/double_cone.png}
            \caption{The equation $h(x, y, z) = x^2 + y^2 - z^2 = 0$ gives rise
            to a double-cone shaped algebraic surface.}
            \label{fig:../pictures/double_cone}
        \end{figure}
        This surface has degree $d = 2$, and since all the terms have the same
        degree the defining polynomial is homogenous, and hence the surface is
        invariant under scaling.  
    \end{exmpl}

    While there is a lot of fun to be had with algebraic surfaces of degree $2$
    there is a limit as to how exciting they are. Things get much more fun when
    we look at surfaces of higher degree. Consider for example the following
    surface of degree $3$.

    \begin{exmpl}[The ding-dong cubic]
        Let $h(x, y, z) = x^2 + y^2 - (1 - z)z^2$. Then the equation $h(x, y,
        z) = 0$ defines a surface of degree $d = 3$ (attributed to the
        $(1-z)z^2$ term). 
        \begin{figure}[h!]
            \centering
            \includegraphics[width=0.3\linewidth]{../pictures/ding_dong.png}
            \caption{The ding-dong cubic given by the equation $h(x, y, z) =
            x^2 + y^2 - (1-z)z^2 = 0$.}
            \label{fig:../pictures/ding_dong}
        \end{figure}
        This surface also exhibits a singularity at the point $(0, 0, 0)$ where
        it passes through itself.
    \end{exmpl}

    Another very interesting  
\clearpage
\printbibliography
\end{document}
